Alleviating the Topology Mismatch Problem in Distributed Overlay Networks

### 2014-02-02 intro.tex
2013.10.20 (Alex) Text reworked and reviewed for introduction and unstructured sections. BibTeX file reformed and debugged (once more :-)

\emph{Peer-to-peer} (\p) networks are self-organizing, distributed systems where
participating nodes, called \emph{peers}, act as both resource providers and
resource consumers, in contrast to the traditional \emph{client-server} model
where nodes undertake specific roles.
For over a decade, \p\ networks have been widely deployed and have
enjoyed immense popularity from Internet communities, primarily because
of the great number of features they offer to distributed applications 
built atop them. 
Such diverse features include:  high availability and robustness,
load-balancing, quality of service, scalability, decentralized administration,
and anonymity.

The peer-to-peer paradigm gave impetus to two ``killer'' applications:
file-sharing and Internet telephony.
The {\sl Napster} file-sharing system was widely acknowledged as the 
``fastest growing Internet application ever'' in $2001$ when it topped
$26$ million users sharing over $80$ million songs.
Other file-sharing applications followed suit, 
including {\sl Gnutella} and {\sl Limewire} enjoying $3$ 
million concurrently-connected peers, as well as 
{\sl BitTorrent} connecting over $150$ million monthly users by January $2012$~\cite{bittorrentusers}.
\p\ telephony saw explosive growth with the advent of {\sl Skype}.
Since its introduction in $2003$,
{\sl Skype} has become extremely popular with more than $650$ million users 
in $2011$~\cite{skypetotalusers} while on January $21$st, $2013$,
the astonishing figure of $50$ million concurrently-online users
was reported~\cite{skypesymusers}.
%%
Apart from the above, though, a startling number of diverse and successful
applications have been built based on \p\ architectures. Some of them include:
distributed search engines~\cite{yaci}, 
distributed data-storage systems~\cite{kbc_oceanstore_2000,bdet_fsdfs_2000,dkkms_cfs_2001,dr_pastutility_2001,abc_farsite_2002,mmfc_ivy_2002,arla,agebh_dks_2003},
Web caches, archives and publishing systems~\cite{ird_squirrel_2002,bags_youserv_2002,wrc_publius_2000,wm_tangler_2001},
messaging and dissemination applications~\cite{threedegrees,icpp08-pd}, 
event-notification infrastructures~\cite{rkcd_scribe_2001,cdkr_scribe_2002,agebh_dks_2003}, 
naming services~\cite{cmm_chorddns_2002}, 
censor-resistant stores~\cite{cswh_freenet_2001} and
lately, even cloud-based platforms~\cite{mgpj_cloudsnap_2011}.

@@

\p\ systems are implemented using \emph{overlay networks}. 
An overlay network, is a virtual system of nodes and logical links
created above an existing network; overlays provide an abstraction that
enables the implementaion of efficient, fully distributed, application-layer
services such as routing messages to destinations that are not known in advance
or offering QoS guarantees (i.e., in content-distribution) over best effort
infrastructures. Overlay nodes communicate through \emph{virtual connections}
each of which may correspond to a path of possibly many physical links 
in the underlying network.
Figure~\ref{figure:overlay} illustrates a simple four-node overlay constructed
over a wide-area network.
%%
\begin{figure}[ht]
\centering
  \includegraphics[scale=0.6]{img/pdf/under-over-lay.pdf}
\caption{An example overlay network.}
\label{figure:overlay}
\end{figure}

The single key issue that determines the efficiency of an overlay network,
is how well the overlay maps to the underlying network topology on which it
``rests''. 
Consider two nodes\footnote{In \p\ networks,
the participating nodes are typically user-PCs operating at the edge of the
Internet.} that are connected with each other via a path of overlay links.
If the application running on the nodes, generates heavy traffic along
the overlay path, it would be beneficial to
construct the overlay topology in a way that the number of 
underlying IP links between these two nodes is minimized.
Should the overlay network be constructed so that
it does not match the underlying topology well, 
the inherent \emph{topology mismatch} gives two
significant problems. First, the performance of the application per se, can be
adversely affected since traffic must flow over a larger,
redundant, number of physical hops resulting in poor user experience 
entailing noticeable latencies or jittering. 
%%
Second, other applications running
on the underlying network infrastructure are also adversely affected.
Studies have shown that highly popular \p\ applications contribute 
the largest portion of the overall 
Internet traffic~\cite{seroiu_analysiscds_2002,sen_analyzep2ptraffic_2004,krp_ispfear_2005}, with some reporting that more than $60$\% of this traffic 
to be \p-related~\cite{cachelogic,ipoque2007,ipoque2009};
it was also projected that this traffic would reach
$8\;$\emph{Petabytes}-per-month by the end of $2012$~\cite{multinteligence}. 
Evidently, this constitutes a major burden for 
Internet Service Providers (ISPs) who must route 
all of this traffic to destinations at the edge of the Internet. 
If the \p\ overlay topology is poorly designed, 
the demand on the Internet's backbone infrastructure may 
substantially increase as traffic might have to flow
``back and forth'' several times between two neighboring ISPs
while trying to travel from the source to its destination node
in the overlay.
Hence, it is very critical that \p\ networks be laid out 
in ways that their topology matches, as closely as possible, the 
underlying IP topology.

>>>>

\emph{Peer-to-peer} (\p) networks are self-organizing, distributed systems where
participating nodes, called \emph{peers}, act as both resource providers and
resource consumers in contrast to the conventional \emph{client-server} model
where nodes undertake specific roles.
For over a decade, \p\ networks have been widely deployed and have
enjoyed immense popularity from Internet communities, primarily because
of the great number of features they offer to distributed applications 
built atop them. 
Such diverse features include:  high availability and robustness,
load-balancing, quality of service, scalability, decentralized administration,
and anonymity.

The peer-to-peer paradigm gave impetus to two ``killer'' applications:
file-sharing and Internet telephony.
The {\sl Napster} file-sharing system was widely acknowledged as the 
``fastest growing Internet application ever'' in $2001$ when it topped
$26$ million users sharing over $80$ million songs.
Other file-sharing applications followed suit, 
including {\sl Gnutella} and {\sl Limewire} enjoying $3$ 
million concurrently-connected peers, as well as 
{\sl BitTorrent} connecting over $150$ million monthly users by January $2012$~\cite{bittorrentusers}.
\p\ telephony saw explosive growth with the advent of {\sl Skype}.
Since its introduction in $2003$,
{\sl Skype} has become extremely popular with more than $650$ million users 
in $2011$~\cite{skypetotalusers} while on January $21$st, $2013$,
the astonishing figure of $50$ million concurrently-online users
was reported~\cite{skypesymusers}.
%%
Apart from the above, a startling number of diverse and successful
applications have been built based on \p\ architectures. Some of them include:
distributed search engines~\cite{yaci}, 
distributed data-storage systems~\cite{kbc_oceanstore_2000,bdet_fsdfs_2000,dkkms_cfs_2001,dr_pastutility_2001,abc_farsite_2002,mmfc_ivy_2002,arla,agebh_dks_2003},
Web caches, archives and publishing systems~\cite{ird_squirrel_2002,bags_youserv_2002,wrc_publius_2000,wm_tangler_2001},
messaging and dissemination applications~\cite{threedegrees,icpp08-pd}, 
event-notification infrastructures~\cite{rkcd_scribe_2001,cdkr_scribe_2002,agebh_dks_2003}, 
naming services~\cite{cmm_chorddns_2002}, 
censor-resistant stores~\cite{cswh_freenet_2001} and
lately, even cloud-based platforms~\cite{mgpj_cloudsnap_2011}.

@@

\p\ systems are implemented using \emph{overlay networks}. 
An overlay network, is a virtual system of nodes and logical links
created above an existing network; overlays provide an abstraction that
enables the implementation of efficient, fully distributed, application-layer
services such as routing messages to destinations that are not known in advance
or offering QoS guarantees (i.e., in content-distribution) over best effort
infrastructures. Overlay nodes communicate through \emph{virtual connections}
each of which may correspond to a path of possibly many physical links 
in the underlying network.
Figure~\ref{figure:overlay} illustrates a simple four-node overlay constructed
over a wide-area network.
%%
\begin{figure}[ht]
\centering
  \includegraphics[scale=0.6]{img/pdf/under-over-lay.pdf}
\caption{An example overlay network.}
\label{figure:overlay}
\end{figure}

The single key issue that determines the efficiency of an overlay network,
is how well the overlay maps to the underlying network topology on which it
``rests''. 
Consider two nodes\footnote{In \p\ networks,
the participating nodes are typically user-PCs operating at the edge of the
Internet.} that are connected with each other via a path of overlay links.
If the application running on the nodes, generates heavy traffic along
the overlay path, it would be beneficial to
construct the overlay topology in a way that the number of 
underlying IP links between these two nodes is minimized.
Should the overlay network be constructed so that
it does not match the underlying topology well, 
the inherent \emph{topology mismatch} gives two
significant problems. First, the performance of the application per se, can be
adversely affected since traffic must flow over a larger,
redundant, number of physical hops resulting in poor user experience 
entailing noticeable latencies or jittering. 
%%
Second, other applications running
on the underlying network infrastructure are also adversely affected.
Studies have shown that highly popular \p\ applications contribute 
the largest portion of the overall 
Internet traffic~\cite{seroiu_analysiscds_2002,sen_analyzep2ptraffic_2004,krp_ispfear_2005}, with some reporting that more than $60$\% of this traffic 
to be \p-related~\cite{cachelogic,ipoque2007,ipoque2009};
it was also projected that this traffic would reach
$8\;$\emph{Petabytes}-per-month by the end of $2012$~\cite{multinteligence}. 
Evidently, this constitutes a major burden for 
Internet Service Providers (ISPs) who must route 
all of this traffic to destinations at the edge of the Internet. 
If the \p\ overlay topology is poorly designed, 
the demand on the Internet's backbone infrastructure may 
substantially increase as traffic might have to flow
``back and forth'' several times between two neighboring ISPs
while trying to travel from the source to its destination node
in the overlay.
Hence, it is critical that \p\ networks be laid out 
in ways that their topology matches, as closely as possible, the 
underlying IP topology.
